---
title: "Biostatistics"
format: revealjs
editor: visual
---

## Dataset Background

Health and mortality indicators extracted directly from the **World Health Organization (WHO)** repository

Corresponding economic data compiled from **United Nations (UN)** databases

Individual data files merged into a unified dataset covering **193 countries**

## Temporal Scope

Spans a 16-year period from **2000 to 2015**

Captures a modern era of rapid global health sector development and significant shifts in human mortality rates

## **Data Cleaning & Quality Control - Not by Us**

Initial visual inspection revealed missing values concentrated primarily in **GDP, Population, and Hepatitis B** metrics.

Data gaps were systematically audited in R using the Missmap command.

Missing observations belonged mostly to less visible/smaller nations (e.g., Vanuatu, Tonga, Togo, Cabo Verde).

**Decision:** These specific countries were excluded from final predictive modeling due to data gaps.

## **Final Dataset Dimensions**

Consists of **2,938 rows** and **22 columns**.

Yields **20 distinct predicting variables** across the 193 countries.

## Summary Statistics

```{r}
library(tidyverse)
library(broom)
library(janitor)
library(skimr)

df_raw <- read.csv("life_expectancy.csv") %>%
  clean_names() %>%
  rename(
    life_exp      = life_expectancy,
    adult_mort    = adult_mortality,
    pct_expend    = percentage_expenditure,
    tot_expend    = total_expenditure,
    under5_deaths = under_five_deaths,
    thinness_1_19 = thinness_1_19_years,
    thinness_5_9  = thinness_5_9_years,
    income_comp   = income_composition_of_resources
  ) %>%
  mutate(status = factor(status, levels = c("Developed", "Developing")))
summary(df_raw)
```

## Missing data

Can be dropped, yields

```{r}
# ── 1.3  Drop rows missing outcome or key predictors; impute remainder --------
df_dropped <- df_raw %>%
  filter(!is.na(life_exp), !is.na(schooling),
         !is.na(income_comp), !is.na(gdp)) %>%
  mutate(across(where(is.numeric),
                ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))
summary(df_dropped)
```

## Missing data

Can be filled by median

```{r}
# VARIATION 2: Grouped median imputation (Specific to Country)
df_med <- df_raw %>%
  # Group by country so mutations happen within each nation's timeline
  group_by(country) %>%
  mutate(across(where(is.numeric), function(x) {
    # Calculate the median for this specific country
    country_median <- median(x, na.rm = TRUE)
    
    # If the country has valid data, use its median. 
    # If it's missing entirely, leave it as NA for the next step.
    if (!is.na(country_median)) {
      ifelse(is.na(x), country_median, x)
    } else {
      x 
    }
  })) %>%
  ungroup() %>% # Always ungroup after group_by modifications
  
  # CRITICAL FALLBACK STEP: 
  # For countries completely missing GDP/Population across all 15 years,
  # impute the global median baseline for that specific year.
  group_by(year) %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  ungroup()
summary(df_med)
```

## Impact of actions
::: {style="overflow-x: auto; overflow-y: auto; max-height: 600px;"}
```{r}
library(tidyverse)
library(gtsummary)

# ------------------------------------------------------------------------------
# 1. COMBINE THE DATASETS
# ------------------------------------------------------------------------------
# We tag each dataset with a new column identifying how it was processed,
# then bind them all together into one massive dataframe.
df_combined <- bind_rows(
  df_raw %>% mutate(Handling_Method = "1. Raw Data (NAs Present)"),
  df_dropped %>% mutate(Handling_Method = "2. Dropped (Only for the key var)"),
  df_med %>% mutate(Handling_Method = "3. Imputed (Country Medians)")
) %>%
  # Select key variables to keep the table clean and readable
  select(Handling_Method, life_exp, schooling, income_comp, gdp, 
         tot_expend, adult_mort, infant_deaths, bmi) %>%
  # Ensure all predictors are numeric to prevent calculation errors
  mutate(across(-Handling_Method, as.numeric))

# ------------------------------------------------------------------------------
# 2. GENERATE THE COMPARISON TABLE
# ------------------------------------------------------------------------------
tbl_comparison <- df_combined %>%
  tbl_summary(
    by = Handling_Method,
    type = list(everything() ~ "continuous"),
    statistic = list(all_continuous() ~ "{mean} ({sd})"),
    digits = all_continuous() ~ 2,
    # 'ifany' explicitly displays a row showing exactly how many NAs are in the 
    # Raw column, while confirming there are 0 in the Dropped and Imputed columns
    missing = "ifany", 
    missing_text = "Missing Values"
  ) %>%
  add_p(
    # Tests for a statistically significant difference across the 3 datasets
    test = list(all_continuous() ~ "kruskal.test"),
    pvalue_fun = ~ style_pvalue(.x, digits = 3)
  ) %>%
  add_overall() %>%
  bold_labels() %>%
  modify_header(label ~ "**Variable**") %>%
  modify_caption("**Table 3. Impact of Missing Data Handling on Variable Distributions**")

# Print the final table
tbl_comparison
```
:::

## Countries
```{r}
# 1. PREP DATA: Isolate unique countries to prevent 16x overcounting
# ------------------------------------------------------------------------------
df_status_counts <- df_raw %>%
  # Keep only one row per country/status combination
  distinct(country, status) %>%
  # Clean up any missing or blank statuses
  filter(!is.na(status) & status != "") %>%
  mutate(status = str_trim(as.character(status)))

df_cal <- df_med %>% #change to df_dropped for dropped NA values
  mutate(
    log_gdp       = log1p(gdp),
    log_population = log1p(population),
    schooling_cat = cut(schooling,
                        breaks = quantile(schooling, c(0, 1/3, 2/3, 1), na.rm = TRUE),
                        labels = c("Low", "Medium", "High"),
                        include.lowest = TRUE),
    income_cat    = cut(income_comp,
                        breaks = quantile(income_comp, c(0, 0.25, 0.75, 1), na.rm = TRUE),
                        labels = c("Low", "Middle", "High"),
                        include.lowest = TRUE),
    life_exp_cat  = cut(life_exp,
                        breaks = c(0, 60, 70, 80, Inf),
                        labels = c("<60", "60-70", "70-80", ">80"),
                        right  = FALSE),
    year_bracket  = cut(year,
                        breaks = c(1999, 2004, 2009, 2015),
                        labels = c("2000-2004", "2005-2009", "2010-2015"),
                        include.lowest = TRUE)
  )
# ------------------------------------------------------------------------------
# 2. GENERATE THE PLOT
# ------------------------------------------------------------------------------
p_status <- ggplot(df_status_counts, aes(x = status, fill = status)) +
  # Draw the bars with a slight transparency (alpha) and crisp black borders
  geom_bar(width = 0.5, color = "black", alpha = 0.85) +
  
  # Automatically calculate the total count and add it as a text label above each bar
  geom_text(stat = "count", aes(label = after_stat(count)), 
            vjust = -0.8, fontface = "bold", size = 5.5) +
  
  # Assign professional, high-contrast colors to the two categories
  scale_fill_manual(values = c("Developed" = "#2C3E50", "Developing" = "#E74C3C")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  # Add clear, descriptive labels
  labs(
    title = "Figure 1. Global Distribution of Country Development Status",
    subtitle = "Total unique countries represented in the WHO dataset",
    x = "Development Status",
    y = "Number of Countries"
  ) +
  
  # Apply a clean, minimalist theme ideal for presentations or Quarto documents
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",                  # Hide the redundant legend
    panel.grid.major.x = element_blank(),      # Remove vertical grid lines for a cleaner look
    plot.title = element_text(face = "bold")
  )

# 3. Render the plot
p_status
```


## Distributions

```{r}
key_vars <- c("life_exp", "schooling", "income_comp", "gdp", "adult_mort", "log_gdp")

# ── Global ggplot theme -------------------------------------------------------
theme_pub <- theme_classic(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(size = 10, colour = "grey40"),
    axis.title      = element_text(face = "bold"),
    legend.position = "bottom",
    strip.text      = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey92", colour = NA),
    panel.grid.major = element_line(colour = "grey92"),
    panel.grid.minor = element_blank()
  )
theme_set(theme_pub)

palette_status <- c("Developed" = "#2166AC", "Developing" = "#D6604D")
# ── Figure 2: Distribution panel ----------------------------------------------
p_hist <- df_cal %>% #used median filled and not the raw
  select(all_of(key_vars)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  mutate(Variable = dplyr::recode(Variable,
    life_exp    = "Life Expectancy",
    schooling   = "Schooling (years)",
    income_comp = "Income Comp. Index",
    gdp     = "GDP per capita",
    log_gdp = "log(GDP per capita)",
    adult_mort  = "Adult Mortality"
  )) %>%
  ggplot(aes(Value)) +
  geom_histogram(bins = 35, fill = "#2166AC", colour = "white", alpha = 0.85) +
  facet_wrap(~ Variable, scales = "free", ncol = 3) +
  labs(title = "Figure 2. Distribution of Key Variables",
       x = NULL, y = "Frequency")
p_hist
```


## Life expectancy

```{r}
# ── Global ggplot theme -------------------------------------------------------
theme_pub <- theme_classic(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(size = 10, colour = "grey40"),
    axis.title      = element_text(face = "bold"),
    legend.position = "bottom",
    strip.text      = element_text(face = "bold"),
    strip.background = element_rect(fill = "grey92", colour = NA),
    panel.grid.major = element_line(colour = "grey92"),
    panel.grid.minor = element_blank()
  )
theme_set(theme_pub)

palette_status <- c("Developed" = "#2166AC", "Developing" = "#D6604D")

p_violin_status <- ggplot(df_dropped, aes(status, life_exp, fill = status)) +
  geom_violin(alpha = 0.35, colour = NA) +
  geom_boxplot(width = 0.22, outlier.shape = 21, outlier.size = 1.5,
               outlier.alpha = 0.5) +
  scale_fill_manual(values = palette_status) +
  labs(title = "Figure 3. Life Expectancy by Country Development Status",
       x = "Development Status", y = "Life Expectancy (years)") +
  guides(fill = "none")
p_violin_status
```

## Average Expenditure
```{r}
#| label: fig-expenditure-trend
#| fig-cap: "Longitudinal trends in health expenditure as a percentage of GDP."
#| warning: false
#| message: false

# 1. PREP DATA: Calculate the yearly average expenditure per group
# ------------------------------------------------------------------------------
df_expend_trend <- df_raw %>%
  # Clean up status and drop rows where expenditure data is entirely missing
  filter(!is.na(status) & status != "", !is.na(tot_expend)) %>%
  mutate(status = str_trim(as.character(status))) %>%
  
  # Group by both year and status to calculate the averages
  group_by(year, status) %>%
  summarise(
    avg_expend = mean(tot_expend, na.rm = TRUE),
    .groups = "drop" # Always drop groups after summarising to prevent downstream bugs
  )

# -------------------------------------------------------------------------
# 2. GENERATE THE PLOT
# ------------------------------------------------------------------------------
p_expend <- ggplot(df_expend_trend, aes(x = year, y = avg_expend, 
                                        color = status, group = status)) +
  
  # Draw the lines and add points for each specific year
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  
  # Apply the same high-contrast color palette for consistency across your slides
  scale_color_manual(values = c("Developed" = "#2C3E50", "Developing" = "#E74C3C")) +
  
  # Force the X-axis to show every 2 or 3 years so it doesn't look cluttered
  scale_x_continuous(breaks = seq(2000, 2015, by = 3)) +
  
  # Add professional labels
  labs(
    title = "Figure 4. Average Health Expenditure by Development Status",
    subtitle = "Tracking expenditure as a percentage of GDP (2000 - 2015)",
    x = "Year",
    y = "Average Health Expenditure (% GDP)",
    color = "Country Status"
  ) +
  
  # Apply a minimalist presentation theme
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank() # Removes the distraction of smaller gridlines
  )

# 3. Render the plot
p_expend
```


## Schooling
```{r}
#| label: fig-schooling-boxplot
#| fig-cap: "Distribution of Schooling Years by Country Development Status"
#| warning: false
#| message: false

library(tidyverse)

# ------------------------------------------------------------------------------
# 1. PREP DATA: Clean status labels and remove missing schooling data
# ------------------------------------------------------------------------------
df_schooling <- df_raw %>%
  filter(!is.na(status) & status != "", !is.na(schooling)) %>%
  mutate(status = str_trim(as.character(status)))

# ------------------------------------------------------------------------------
# 2. GENERATE THE BOXPLOT
# ------------------------------------------------------------------------------
p_school_box <- ggplot(df_schooling, aes(x = status, y = schooling, fill = status)) +
  
  # Draw the boxplot. We make it slightly narrower (width) for elegance, 
  # and style the outlier dots so they look deliberate rather than messy.
  geom_boxplot(width = 0.4, alpha = 0.85, color = "black",
               outlier.shape = 21, outlier.size = 2, outlier.alpha = 0.5) +
  
  # Keep the high-contrast color palette consistent with your other charts
  scale_fill_manual(values = c("Developed" = "#2C3E50", "Developing" = "#E74C3C")) +
  
  # Add professional labels
  labs(
    title = "Figure 5. Distribution of Schooling by Development Status",
    subtitle = "Comparing total years of schooling across all available data (2000-2015)",
    x = "Development Status",
    y = "Schooling (Years)"
  ) +
  
  # Apply a clean presentation theme
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",              # The X-axis handles the labels, no legend needed
    panel.grid.major.x = element_blank(),  # Remove vertical lines behind the boxes
    plot.title = element_text(face = "bold")
  )

# 3. Render the plot
p_school_box
```

## But..
```{r}
#I dont like this one, prob skip, too crowded
#| label: fig-schooling-panel
#| fig-cap: "Year-by-year distribution of Schooling by Country Development Status"
#| warning: false
#| message: false

# 1. PREP DATA
# ------------------------------------------------------------------------------
df_schooling <- df_raw %>%
  filter(!is.na(status) & status != "", !is.na(schooling)) %>%
  mutate(status = str_trim(as.character(status)))

# ------------------------------------------------------------------------------
# 2. GENERATE THE FACETED PLOT
# ------------------------------------------------------------------------------
p_school_panel <- ggplot(df_schooling, aes(x = status, y = schooling, fill = status)) +
  
  # Draw the boxplot. Outlier sizes are reduced to prevent clutter in small panels
  geom_boxplot(width = 0.6, alpha = 0.85, color = "black",
               outlier.shape = 21, outlier.size = 1, outlier.alpha = 0.5) +
  
  # THE MAGIC LINE: This splits the plot into 16 panels, one for each year
  facet_wrap(~ year, nrow = 4) +
  
  scale_fill_manual(values = c("Developed" = "#2C3E50", "Developing" = "#E74C3C")) +
  
  labs(
    title = "Figure 6. Evolution of Schooling by Development Status",
    subtitle = "Year-by-year tracking of educational attainment (2000-2015)",
    x = NULL, # X-axis title removed to save precious vertical space on the slide
    y = "Schooling (Years)"
  ) +
  
  # Apply theme optimized for faceted presentation slides
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "none",
    
    # Angle the X-axis text slightly so "Developed" and "Developing" don't crash into each other
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8), 
    
    # Style the panel headers (the years) to make them pop
    strip.background = element_rect(fill = "gray90", color = "white"),
    strip.text = element_text(face = "bold", size = 10),
    
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold", size = 14)
  )

# 3. Render the plot
p_school_panel
```


## BMI

```{r}
#| label: fig-bmi-extremes
#| fig-cap: "Countries with the highest and lowest mean BMI (2015)"
#| warning: false
#| message: false


# 1. PREP DATA: Isolate 2015 and extract the Top/Bottom 10
# ------------------------------------------------------------------------------
df_bmi_2015 <- df_raw %>%
  # Filter for the most recent year to get an accurate modern snapshot
  filter(year == 2015, !is.na(bmi)) %>%
  select(country, bmi)

# Extract the 10 highest and 10 lowest countries, tagging them for colors later
top_10 <- df_bmi_2015 %>% 
  slice_max(bmi, n = 10) %>% 
  mutate(Category = "Highest BMI")

bottom_10 <- df_bmi_2015 %>% 
  slice_min(bmi, n = 10) %>% 
  mutate(Category = "Lowest BMI")

# Combine them and lock the factor levels so the chart sorts cleanly from top to bottom
df_bmi_extremes <- bind_rows(top_10, bottom_10) %>%
  mutate(country = fct_reorder(country, bmi))


# 2. GENERATE THE PLOT
# ------------------------------------------------------------------------------
p_bmi <- ggplot(df_bmi_extremes, aes(x = bmi, y = country, fill = Category)) +
  
  # Draw horizontal bars
  geom_col(color = "black", alpha = 0.85, width = 0.7) +
  
  # Automatically stamp the exact BMI number onto the end of each bar for readability
  geom_text(aes(label = round(bmi, 1)), 
            hjust = -0.2, size = 4, fontface = "bold") +
  
  # Use your established presentation color palette
  scale_fill_manual(values = c("Highest BMI" = "#2C3E50", "Lowest BMI" = "#E74C3C")) +
  
  # Expand the X-axis slightly so the text labels don't get cut off at the edge
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  
  labs(
    title = "Figure 6. Global BMI Extremes",
    subtitle = "Top 10 and Bottom 10 countries by Mean BMI (2015)",
    x = "Mean Body Mass Index (BMI)",
    y = NULL # Removes the redundant "country" axis label
  ) +
  
  # Apply a minimalist theme
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",              # Hide redundant legend
    panel.grid.major.y = element_blank(),  # Remove horizontal lines behind the bars
    plot.title = element_text(face = "bold"),
    axis.text.y = element_text(face = "bold")
  )

# 3. Render the plot
p_bmi
```

## Global Life Expectancy 2000
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-align: center
library(tidyverse)
library(maps)
library(viridis) 

# 1. PREP THE WHO DATA (Filter years and fix country names)
# ----------------------------------------------------------------------------
df_map_data <- df_raw %>%
  filter(year %in% c(2000, 2005, 2010, 2015), !is.na(life_exp)) %>%
  select(country, year, life_exp) %>%
  # Translate WHO formal names to match the 'maps' package dictionary
  mutate(country = case_match(country,
    "United States of America"                             ~ "USA",
    "United Kingdom of Great Britain and Northern Ireland" ~ "UK",
    "Russian Federation"                                   ~ "Russia",
    "Republic of Korea"                                    ~ "South Korea",
    "Democratic People's Republic of Korea"                ~ "North Korea",
    "Iran (Islamic Republic of)"                           ~ "Iran",
    "Bolivia (Plurinational State of)"                     ~ "Bolivia",
    "Venezuela (Bolivarian Republic of)"                   ~ "Venezuela",
    "Viet Nam"                                             ~ "Vietnam",
    "Syrian Arab Republic"                                 ~ "Syria",
    "United Republic of Tanzania"                          ~ "Tanzania",
    "Côte d'Ivoire"                                        ~ "Ivory Coast",
    "Lao People's Democratic Republic"                     ~ "Laos",
    "Congo"                                                ~ "Republic of Congo",
    .default = as.character(country)
  ))

# 2. PREP THE MAP BASE (Expand for all 4 years)
# ------------------------------------------------------------------------------
# Extract the base world map
world_base <- map_data("world") %>%
  filter(region != "Antarctica") # Drop Antarctica to save space

# To facet polygons properly without dropping countries that lack data in a specific 
# year, we cross the base map with our target years.
world_expanded <- expand_grid(
  world_base,
  year = c(2000, 2005, 2010, 2015)
)

# Join our cleaned WHO data to the expanded map coordinates
map_final <- world_expanded %>%
  left_join(df_map_data, by = c("region" = "country", "year" = "year"))

# map 1
map_2000 <- map_final %>% filter(year == 2000)

ggplot(map_2000, aes(x = long, y = lat, group = group, fill = life_exp)) +
  geom_polygon(color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(option = "magma", direction = -1, na.value = "gray85", 
                       name = "Life Expectancy", limits = c(40, 85)) +
  theme_void(base_size = 14) + 
  theme(legend.position = "bottom", legend.key.width = unit(2, "cm"))
```

## Global Life Expectancy 2005
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-align: center

map_2005 <- map_final %>% filter(year == 2005)

ggplot(map_2005, aes(x = long, y = lat, group = group, fill = life_exp)) +
  geom_polygon(color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(option = "magma", direction = -1, na.value = "gray85", 
                       name = "Life Expectancy", limits = c(40, 85)) +
  theme_void(base_size = 14) + 
  theme(legend.position = "bottom", legend.key.width = unit(2, "cm"))
```

## Global Life Expectancy 2010
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-align: center

map_2010 <- map_final %>% filter(year == 2010)

ggplot(map_2010, aes(x = long, y = lat, group = group, fill = life_exp)) +
  geom_polygon(color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(option = "magma", direction = -1, na.value = "gray85", 
                       name = "Life Expectancy", limits = c(40, 85)) +
  theme_void(base_size = 14) + 
  theme(legend.position = "bottom", legend.key.width = unit(2, "cm"))
```

## Global Life Expectancy 2015
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-align: center

map_2015 <- map_final %>% filter(year == 2015)

ggplot(map_2015, aes(x = long, y = lat, group = group, fill = life_exp)) +
  geom_polygon(color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(option = "magma", direction = -1, na.value = "gray85", 
                       name = "Life Expectancy", limits = c(40, 85)) +
  theme_void(base_size = 14) + 
  theme(legend.position = "bottom", legend.key.width = unit(2, "cm"))
```



## Combined
```{r}
# 3. GENERATE THE STATIC PLOT
# ------------------------------------------------------------------------------
p_world_map <- ggplot(map_final, aes(x = long, y = lat, group = group, fill = life_exp)) +
  
  # Draw the country polygons. NA values (missing data) are colored light gray.
  geom_polygon(color = "white", linewidth = 0.1) +
  
  # Split the map into our 4 chosen years
  facet_wrap(~ year, ncol = 2) +
  
  # Apply a high-contrast, colorblind-friendly color gradient (viridis)
  scale_fill_viridis_c(
    option = "magma",         # "magma" transitions beautifully from dark purple to bright yellow
    direction = -1,           # Flips it so dark = lower life expectancy, bright = higher
    na.value = "gray85",      # Color for countries missing from the dataset
    name = "Life Expectancy"
  ) +
  
  labs(
    title = "Figure 7. Global Life Expectancy (2000 - 2015)",
    subtitle = "Tracking longitudinal improvements across a 15-year span"
  ) +
  
  # theme_void() strips away the latitude/longitude axes and gridlines
  theme_void(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(2, "cm"),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, margin = margin(b = 15)),
    strip.text = element_text(face = "bold", size = 12, margin = margin(b = 5))
  )

# Render the plot
p_world_map
```


## Is the data normally distributed?
::: {style="overflow-x: auto; overflow-y: auto; max-height: 600px;"}
```{r}
library(gt)
key_vars <- c("life_exp", "schooling", "income_comp", "gdp", "adult_mort", "log_gdp") #can use log_gdp

tbl2 <- purrr::map(key_vars, function(v) {
  
  x <- as.numeric(na.omit(df_cal[[v]])) 
  
  # 1. THE FIX: Safety check for minimum sample size required by shapiro.test
  if (length(x) >= 3) {
    sw <- shapiro.test(sample(x, min(5000, length(x))))
    sw_w   <- round(sw$statistic, 4)
    sw_p   <- round(sw$p.value, 4)
    normal <- ifelse(sw$p.value > 0.05, "Yes", "No")
  } else {
    # If N < 3, gracefully return NAs instead of crashing
    sw_w   <- NA
    sw_p   <- NA
    normal <- "N/A"
  }
  
  tibble(
    Variable  = v,
    N         = length(x), 
    Mean      = round(mean(x, na.rm = TRUE), 2),
    SD        = round(sd(x, na.rm = TRUE), 2),
    Skewness  = round(moments::skewness(x, na.rm = TRUE), 3),
    Kurtosis  = round(moments::kurtosis(x, na.rm = TRUE), 3),
    SW_W      = sw_w,
    SW_p      = sw_p,
    Normal    = normal
  )
  
}) %>% 
  purrr::list_rbind() %>%   
  mutate(
    Variable = dplyr::recode_values(Variable,
                                    "life_exp"    ~ "Life Expectancy", 
                                    "schooling"   ~ "Schooling",
                                    "income_comp" ~ "Income Comp. Index",      
                                    "gdp"     ~ "GDP",
                                    "log_gdp" ~ "log GDP",
                                    "adult_mort"  ~ "Adult Mortality",         
                                    default = Variable
    )
  )
tbl2_gt <- tbl2 %>%
  gt() %>%
  tab_header(title = "Table 2. Normality Assessment of Key Continuous Variables") %>%
  tab_spanner(label = "Descriptive", columns = c(N, Mean, SD, Skewness, Kurtosis)) %>%
  tab_spanner(label = "Shapiro-Wilk", columns = c(SW_W, SW_p, Normal)) %>%
  cols_label(SW_W = "W", SW_p = "p-value", Normal = "Normal?") %>%
  fmt_number(columns = c(Mean, SD, Skewness, Kurtosis, SW_W), decimals = 3) %>%
  fmt_number(columns = SW_p, decimals = 4) %>%
  tab_style(
    style = cell_text(color = "red"),
    locations = cells_body(columns = Normal, rows = Normal == "No")
  ) %>%
  tab_footnote("Shapiro-Wilk W; p < 0.05 indicates departure from normality. Sample drawn (n=5000) for large datasets.")
tbl2_gt
```
:::

## Combined

```{r}
#| label: fig-density-panel
#| fig-cap: "Density distribution of continuous variables by country status"
#| warning: false
#| message: false

# 1. PREP THE DATA (The Wide-to-Long Transformation)
# ------------------------------------------------------------------------------
df_long <- df_cal %>%
  # Keep only valid status labels
  filter(!is.na(status) & status != "") %>%
  mutate(status = str_trim(as.character(status))) %>%
  
  # Select ONLY the grouping column (status) and the continuous variables you want to plot
  select(status, adult_mort, bmi, gdp, infant_deaths, life_exp, tot_expend, schooling) %>%
  
  # Melt all numeric columns into two columns ("variable" and "value")
  pivot_longer(
    cols = -status,        # Pivot everything EXCEPT the status column
    names_to = "variable", # Put the column names here
    values_to = "value"    # Put the actual numbers here
  ) %>%
  
  # Drop missing values to prevent plotting errors
  filter(!is.na(value)) %>%
  
  # Clean up the variable names so the panel headers look professional
  mutate(variable = case_match(variable,
    "adult_mort"    ~ "Adult Mortality",
    "bmi"           ~ "BMI",
    "gdp"           ~ "GDP",
    "infant_deaths" ~ "Infant Deaths",
    "life_exp"      ~ "Life Expectancy",
    "tot_expend"    ~ "Total Expenditure",
    "schooling"     ~ "Schooling",
    .default = variable
  ))


# 2. GENERATE THE DENSITY PLOTS
# ------------------------------------------------------------------------------
p_density <- ggplot(df_long, aes(x = value, fill = status)) +
  
  # Draw the smooth, overlapping curves
  geom_density(alpha = 0.5, color = "black", linewidth = 0.3) +
  
  # Facet by the new 'variable' column we created in step 1
  facet_wrap(~ variable, scales = "free") +
  
  scale_fill_manual(values = c("Developed" = "#2C3E50", "Developing" = "#E74C3C")) +
  
  labs(
    title = "Figure 8. Distribution Overlap by Development Status",
    subtitle = "Smoothed density curves (Area under each curve = 100%)",
    x = NULL, 
    y = "Density"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    plot.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold", size = 12), 
    panel.grid.minor = element_blank()
  )

# 3. Render the plot
p_density
```


## GDP and Life Expectancy

See correlation

## Descriptive Summary
::: {style="overflow-x: auto; overflow-y: auto; max-height: 600px;"}
```{r}
df_years <- df_cal %>%
  # Select milestone years to keep the table readable
  filter(year %in% c(2000, 2007, 2015)) %>%
  
  # ERROR FIX: Strip NAs and force 'status' into a clean character string 
  # to destroy any phantom factor levels causing the 2-level error
  filter(!is.na(status) & status != "") %>%
  mutate(status = str_trim(as.character(status))) %>%
  
  # Force predictors to numeric to prevent the Chi-Square bug
  mutate(across(-c(country, status, year), as.numeric)) %>%
  
  # Select final columns
  select(year, status, life_exp, schooling, income_comp, gdp,
         tot_expend, adult_mort, infant_deaths, hiv_aids, bmi, alcohol)


# 2. GENERATE THE STRATIFIED YEAR-BY-YEAR TABLE -- better------
# ------------------------------------------------------------------------------
tbl_yearly <- df_years %>%
  # tbl_strata splits the table into side-by-side columns based on the 'year'
  tbl_strata(
    strata = year,
    .tbl_fun = ~ .x %>%
      tbl_summary(
        by = status,
        type = list(everything() ~ "continuous"), 
        statistic = list(all_continuous() ~ "{mean} ({sd})"),
        digits  = all_continuous() ~ 2,
        label   = list(
          life_exp      ~ "Life Expectancy (years)",
          schooling     ~ "Schooling (years)",
          income_comp   ~ "Income Composition Index",
          gdp           ~ "GDP per capita",
          tot_expend    ~ "Health Expenditure (% GDP)",
          adult_mort    ~ "Adult Mortality (per 1,000)",
          infant_deaths ~ "Infant Deaths (per 1,000)",
          hiv_aids      ~ "HIV/AIDS Deaths (per 1,000)",
          bmi           ~ "Mean BMI",
          alcohol       ~ "Alcohol Consumption (L/capita)"
        ),
        missing = "no"
      ) %>%
      add_p(
        test = list(all_continuous() ~ "wilcox.test"),
        pvalue_fun = ~ style_pvalue(.x, digits = 3)
      ) %>%
      modify_header(label ~ "**Variable**")
  ) %>%
  # Add a clean caption to the overall stratified table
  modify_caption("**Table 2.** Longitudinal characteristics by country development status (2000 vs 2007 vs 2015)")

# Print the table
tbl_yearly
```
:::

## Logistic Regression

Since we do not have binary variables to analyse, we chose to confirm that indeed higher life expectancy (\>70 years) can be achieved by immunisation, schooling and higher GDP

What is the probability that a country achieves a high life expectancy (greater than 70 years), given its immunization coverage, economic baseline, and educational attainment?

## Model Specifications

$$\log\left(\frac{P(Y=1)}{1-P(Y=1)}\right)=\beta_0+ \beta_1 \mathrm{Polio}+ \beta_2 \mathrm{Diphtheria}+ \beta_3 \log(\mathrm{GDP})+ \cdots$$

|  |  |  |
|---------------------|-----------------------|----------------------------|
| **Component** | **Variable in Dataset** | **Role** |
| **Outcome (**$Y$) | high_life_exp | 1 if life_exp $\ge$ 70, 0 if otherwise |
| **Predictor (**$X_1$) | polio | Immunization coverage (%) |
| **Predictor (**$X_2$) | diphtheria | Immunization coverage (%) |
| **Predictor (**$X_3$) | gdp (log-transformed) | Economic determinant |
| **Predictor (**$X_4$) | schooling | Social determinant (Years) |
| **Confounder (**$X_5$) | status | Categorical adjustment (Developed/Developing) |

::: {style="overflow-x: auto; overflow-y: auto; max-height: 600px;"}
```{r}

# Install the panel data package if you don't have it
# install.packages("plm")

library(plm)
library(broom)

# 1. Declare your dataset as Panel Data
# This tells R that the data is clustered by 'country' and progresses by 'year'
pdata <- pdata.frame(df_cal, index = c("country", "year"))

# 2. Run a Two-Way Fixed Effects Regression
# 'effect = "twoways"' controls for BOTH unobserved country traits AND global year shocks
panel_model <- plm(
  life_exp ~ log(gdp) + schooling + adult_mort + infant_deaths + polio, 
  data = pdata, 
  model = "within",   # "within" is the econometric term for Fixed Effects
  effect = "twoways"  
)

# 3. View the Results
panel_model

# 4. Clean output for presentation
tidy(panel_model, conf.int = TRUE)
```
:::

## How do these compare
::: {style="overflow-x: auto; overflow-y: auto; max-height: 600px;"}
```{r}
# Install packages if you don't have them
# install.packages(c("plm", "modelsummary", "ggplot2"))

library(tidyverse)
library(plm)
library(modelsummary)
library(ggplot2)

# ------------------------------------------------------------------------------
# 1. PREPARE THE DATA
# ------------------------------------------------------------------------------
# Ensure the data is declared as a panel dataset for the fixed effects model
pdata <- pdata.frame(df_cal, index = c("country", "year"))

# ------------------------------------------------------------------------------
# 2. RUN THE MODELS
# ------------------------------------------------------------------------------
# Model 1: Naive Standard Regression (Ignores country and year grouping)
model_naive <- lm(
  life_exp ~ log(gdp) + schooling + adult_mort + infant_deaths + polio, 
  data = df_cal
)

# Model 2: Two-Way Fixed Effects (Controls for specific countries and global years)
model_twfe <- plm(
  life_exp ~ log(gdp) + schooling + adult_mort + infant_deaths + polio, 
  data = pdata, 
  model = "within", 
  effect = "twoways"
)

# Bundle models into a named list for comparison
models_to_compare <- list(
  "Naive OLS (Standard)" = model_naive,
  "Two-Way Fixed Effects" = model_twfe
)

# ------------------------------------------------------------------------------
# 3. TABULAR COMPARISON (Side-by-Side Table)
# ------------------------------------------------------------------------------
# This generates a beautifully formatted table showing coefficients and standard errors
comparison_table <- modelsummary(
  models_to_compare,
  output = "gt",
  stars = TRUE, # Adds significance stars (* p < 0.05, ** p < 0.01, etc.)
  title = "Table 4. Model Comparison: Standard OLS vs Fixed Effects",
  coef_rename = c("log(gdp)" = "log(GDP)",
                  "schooling" = "Schooling (Years)",
                  "adult_mort" = "Adult Mortality",
                  "infant_deaths" = "Infant Deaths",
                  "polio" = "Polio Immunization (%)")
)

comparison_table

```
:::

# Comparison...
```{r}
# 4. VISUAL COMPARISON (Coefficient Plot)
# ------------------------------------------------------------------------------
# This plots the coefficients and their 95% confidence intervals side-by-side
# We omit the Intercept because Fixed Effects models absorb it, making it uncomparable.
visual_plot <- modelplot(
  models_to_compare, 
  coef_omit = "Intercept",
  coef_rename = c("log(gdp)" = "log(GDP)",
                  "schooling" = "Schooling",
                  "adult_mort" = "Adult Mortality",
                  "infant_deaths" = "Infant Deaths",
                  "polio" = "Polio Immunization")
) +
  # Add ggplot2 styling for a clean, professional look
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
  theme_minimal() +
  scale_color_manual(values = c("Naive OLS (Standard)" = "#E74C3C", 
                                "Two-Way Fixed Effects" = "#2C3E50")) +
  labs(
    title = "Coefficient Comparison: Standard vs. Fixed Effects Model",
    subtitle = "Points represent the estimated effect on Life Expectancy. Bars are 95% CIs.",
    x = "Estimated Coefficient (Effect on Life Expectancy)",
    y = "Predictor Variable",
    color = "Model Type"
  ) +
  theme(
    legend.position = "bottom",
    axis.text.y = element_text(face = "bold", size = 11)
  )

visual_plot
```

